Revisiting Multi-Domain Machine Translation

### 2020-04-15 2020/coling2020/multidomain.tex
review first sections

\documentclass[11pt]{article}
\usepackage{coling2020}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[draft]{todo}
%\setlength\titlebox{5cm}
%\colingfinalcopy % Uncomment this line for the final submission

@@

\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{\textbf{#1}}}

@@

Data-based Machine Translation (MT), whether statistical or neural, rests on well understood machine learning principles. Given a training sample of matched source-target sentence pairs $(\src,\trg)$\jcDone{shoulnt trg be e as below ?} drawn from an underlying distribution $\mathcal{D}$, a model parameterized by $\theta$ (here a translation function $f_{\theta}$) is trained by minimizing the empirical expectation of a loss function $\ell(f_\theta(\src), \trg)$. This approach ensures that the expectation of the translation loss will remain low when translating new sentences \emph{drawn from the same distribution}.

Owing to the great variability of language data, this ideal situation is rarely met in practice, warranting the study of an alternative scenario, where the test distribution $\mathcal{D'}$ differs from $\mathcal{D}$. In this setting, \emph{domain adaptation} (DA) methods are required. DA has a long history in Machine Learning in general (eg.\ \cite{Shimodaira00improving,Ben-David09atheory}) and in NLP in particular (eg.\ \cite{Daume06domain,Blitzer07domain,Jiang07instance}), and various techniques exist to handle both the situations where a (small) training sample drawn from $\mathcal{D'}$ is available in training, or where only samples of source-side (or target-side) sentences are available (see \cite{Foster07mixture,Bertoldi09domain,Axelrod11domain} for early proposals for statistical MT, or \cite{Chu18asurvey} for a recent survey of DA in the context of Neural MT).
% In a nutshell, DA attempts to compensate for the distribution mismatch either by biasing the training sample to match ${\mathcal{D'}$ (eg.\ using resampling or instance-weighting strategies), or by adapting the

A seemingly related problem is \emph{multi-domain (MD) machine translation} \cite{Hassan17neural,Farajian17multidomain,Kobus17domaincontrol,Zeng18multidomain,Pham19generic} where a single system has to be trained  and tested with data from multiple domains. As pointed out eg.\ by \newcite{Dredze08online}, this setting borrows much from DA, but also from multi-task learning \cite{Caruana97multitask}. If the intuitions behind MD may seem simple, the exact specifications of MDMT systems are rarely spelled out, which makes their evaluation difficult: for instance, should they perform well when the test data is distributed like the train data, when it is equally distributed across training domains or when the test distribution is unknown? should MDMT also be robust to new domains? How should they handle domain labeling errors?

@@

\section{Requirements of multi-domain MT \label{sec:requirements}}
In this section, we try to recap the main reasons for considering a multi-domain scenario and discuss their implications in terms of performance evaluation.
\fyTodo{Relation to theory, discussion of worst case scenario}

\subsection{MDMT: a tentative definition}
Informally, a multi-domain system should be able to accurately translate heterogeneous sources of data. This loose definition correspond to several practical situations:
\begin{itemize}
\item the more common situation is when both train and test documents (or sentences) are domain-tagged, where each domain corresponds to a specific source of data, and differ from the other sources in terms of textual genre, thematic content \cite{Zhang16topicinformed}, register, style \cite{Niu18multitask}, etc. It is usually assumed that the set of test domains is the same as the set of training domains -- when this is not the case, we would have zero-shot domain adaptation, similar to what is studied in multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Platanios18contextual}. This is the more general setting, already proposed in \cite{Dredze09multidomain}.\fyTodo{do we need a name here: such as supervised MDMT ? and mostly supervised MDMT for the second case ?}

@@

\subsection{Reasons for using MDMT? \label{ssec:whymdmt}}
The motivations for considering MDMT are many. Looking at the literature, the main arguments in support of MDMT seem to be the following:
\begin{enumerate}
\item[[M1\hspace{-0.5em}]] computational / practical arguments. A first motivation for MDMT is practical: when faced with inputs that are potentially from multiple domains, it is easier and computationally cheaper to develop only one system instead of having to optimize (and maintain) multiple MT engines \cite{Sennrich13multidomain,Farajian17neural}. Such arguments are all the more convincing that the number of domains considered is large, which is not always the case: for instance \cite{Britz17mixing,Zeng18multidomain,Jiang19multidomain} only consider 2-4 domains in their experiments.

\item[[M2\hspace{-0.5em}]] statistical arguments. The training data available for each domain is very unevenly distributed, and domain trained or adapted on small datasets will yield large estimator variance. Building mix-domain systems is likely to reduce this variance, at the expense of a larger statistical bias \cite{Clark12onesystem}. Under this view, MDMT would be especially useful for domains with little training data. This is what is observed for multilingual MT from English: an increase for under-resource languages due to positive transfer, at the cost of a decrease for well-resourced languages \cite{Arivazhagan19massively}. A slightly distinct line of reasoning is that mixing domains has a positive regularization effect for all domains. By introducing variability in training, it prevents domain adaptation from overfitting the available adaptation data and could help improve generalization even for well-resourced domains.

\item[[M3\hspace{-0.5em}]] linguistic arguments. A related line of reasoning contends that domain specificities are mostly expressed lexically and will primarily affect content words or multi-word expressions; function words, on the other hand, are domain agnostic and remain semantically stable accross domain. A MDMT system should simultaneously learn lexical domain peculiarities, and leverage cross-domain similarities to improve the representation of generic contexts and words \cite{Zeng18multidomain,Pham19generic}. It is expected that the MDMT scenario should be more profitable when the domain mix includes domains that are closely related and which can share more information. \fyTodo{polysemy}
\end{enumerate}

These various reasons for adopting MDMT are not mutually exclusive. Each however yields specific expectations with respect to the performance of this approach: if the motivation is primarily computational, then a drop in MT quality with respect to fine-tuning each domain might be acceptable; if it is to improve statistical estimation, then the hope will be that MDMT will improve, at least for some domains, if not all, over individually trained systems.

\cite[eq. (3)]{Wang20general} defines the loss of MDMT as the (macro) average of domain loss - all domains are equally important.

@@

\section{Challenging multi-domain translation systems}
In this section, we discuss the operational requirements that can be expected from a true multi-domain system, as well as ways to evaluate how much these requirements are met in practice.

\subsection{Multi-domain systems should be effective \label{ssec:effective}}
% \subsubsection{Performance}
The first set of requirements for MDMT systems is that they should perform well in the face of mix-domain data. These yields the following desired properties.

\paragraph{[P1-LAB]}\fyTodo{Decide naming scheme} A first expectation for MDMT is that they perform better that the obvious baseline which disregards domain labels. % Testing that MDMT actually benefit from domain tags is simple, notwithstanding the previous remarks above regarding evaluation procedures.

\paragraph{[P2-TUN]} Additionally, following motivation~(M2) above, one would expect that MDMT improves over fine-tuning, at least in domains where data is scarce, or in situations where several domains are close. To evaluate this, we perform two measurements, using a real as well as an artificial scenario. In the real scenario, we simply compare the performance of MDMT and fine-tuning for domains of varying sizes, expecting a larger gain for smaller domains (property [P2.1-TUN]). In the artificial scenario, we split a single training domain in two parts which are considered as distinct in training. The expectation here is that a MDMT should yield a clear gain in the both pseudo sub-domains, which should benefit from the supplementary amount of relevant training (this is property [P2.2-TUN]). In this situation, MDMT should even outperform fine-tuning on either of the pseudo sub-domain.
\jcDone{I dont understand the prediction... why running MDMT on 2 artificial subdomains should show  gains over fine-tunning?}\fyTodo{Clearer now ?}

\subsection{Robustness with respect to domain definition \label{ssec:robusness}}
A second set of requirements is related to the definition of a domain. As repeatedly pointed out in the literature, training corpora for MT are often collected opportunistically and the view that each corpora constitutes a single domain is often a gross approximation.\footnote{Two of our own domains are actually composed of several subcorpora (IT and MED), see details in section~\ref{ssec:corpora}.} MDMT should aim to make the best of the available data and be robust to domain assignments. To challenge these requirements we propose the evaluate the following properties.

\paragraph{[P3-HET]}
The notion of a domain is often a fragile one and an effective MDMT system should be able to discover not only when cross-domain sharing is most useful (cf.\ property [P2.2-TUN]) \jcDone{what is P2.2?}, but also when intra-domain heterogeneity is hurting. Requirement [P3-HET] is tested by artificially conjoining two separate domains in one during training. The loss in performance with respect to the baseline (correct domain tags) should then be as small as possible.

\paragraph{[P4-ERR]}
A MDMT should perform best when the true domain tag is known, but deteriorate gracefully in the presence of domain tag errors. This is in sharp contrast with what is commonly observed with fine-tuning, where giving the wrong domain tag can yield ``catastrophic'' drop in performance. This requirement is measured by decoding the test set for domain A with all other domain tags, and report average drop in performance.

\paragraph{[P5-UNK]}
A related situation occurs when the domain of a test document is unknown. Several alternatives are in order: (a) automatically assigning one of the domain tag for decoding should not be much worse than using the true domain tag, assuming this domain is seen in training. For test documents from other domains, the same procedure can apply, and should outperform the use of a single mix-domain system. \fyTodo{Systems react to unknown domains}These requirements are respectively labelled [P5.1] and [P5.2]\fyTodo{Not so sure of these ones}.

>>>>

\documentclass[11pt]{article}
\usepackage{coling2020}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[draft]{todo}
\usepackage{xspace}
\usepackage{amsmath,amsfonts,amssymb}
%\setlength\titlebox{5cm}
%\colingfinalcopy % Uncomment this line for the final submission

@@

\newcommand{\src}{\ensuremath{\mathbf{f}}} % source sentence
\newcommand{\trg}{\ensuremath{\mathbf{e}}} % target sentence
\newcommand{\domain}[1]{\texttt{\textsc{#1}}}
\newcommand{\system}[1]{\texttt{\textbf{#1}}}
\newcommand{\vlambda}{\ensuremath{\boldsymbol\lambda}\xspace} % parameters vector for a distribution

@@

Data-based Machine Translation (MT), whether statistical or neural, rests on well understood machine learning principles. Given a training sample of matched source-target sentence pairs $(\src,\trg)$\jcDone{shoulnt trg be e as below ?} drawn from an underlying distribution $\mathcal{D}$, a model parameterized by $\theta$ (here a translation function $f_{\theta}$) is trained by minimizing the empirical expectation of a loss function $\ell(f_\theta(\src), \trg)$. This approach ensures that the expectation of the translation loss will remain low when translating new sentences drawn from the same distribution.

Owing to the great variability of language data, this ideal situation is rarely met in practice, warranting the study of an alternative scenario, where the test distribution $\mathcal{D'}$ differs from $\mathcal{D}$. In this setting, \emph{domain adaptation} (DA) methods are required. DA has a long history in Machine Learning in general (eg.\ \cite{Shimodaira00improving,Ben-David09atheory})\fyTodo{cite non covariate shift} and in NLP in particular (eg.\ \cite{Daume06domain,Blitzer07domain,Jiang07instance}), and various techniques exist to handle both the situations where a (small) training sample drawn from $\mathcal{D'}$ is available in training, or where only samples of source-side (or target-side) sentences are available (see \cite{Foster07mixture,Bertoldi09domain,Axelrod11domain} for proposals from the statistical MT era, or \cite{Chu18asurvey} for a recent survey of DA in the context of Neural MT).
% In a nutshell, DA attempts to compensate for the distribution mismatch either by biasing the training sample to match ${\mathcal{D'}$ (eg.\ using resampling or instance-weighting strategies), or by adapting the

A seemingly related problem, that has important practical implications, is multi-domain (MD) machine translation \cite{Hassan17neural,Farajian17multidomain,Kobus17domaincontrol,Zeng18multidomain,Pham19generic} where one single system has to be trained and tested with data from multiple domains. As pointed out both by \newcite{Dredze08online} and \cite{Finkel09hierarchical}, this setting borrows much from DA, but also from multi-task learning \cite{Caruana97multitask,Yang15unified}; note that it also share commonalties with ensemble learning \cite{XXX}.\fyTodo{Missing ref} If the intuitions behind MDMT are simple, the exact specifications of MDMT systems are rarely spelled out: for instance, should they perform well when the test data is distributed like the train data, when it is equally distributed across training domains or when the test distribution is unknown? Should MDMT also be robust to new domains? How should they handle domain labeling errors? etc. % , which makes their evaluation fragile

@@

\section{Requirements of multi-domain MT \label{sec:requirements}}
In this section, we try to recap the main reasons for considering a multi-domain scenario and discuss their implications in terms of performance evaluation.
\fyTodo{Relation to theory, discussion of worst case scenario}
\fyTodo{Insist on the work of Dredze}

\subsection{Formalizing multi-domain learning \label{ssec:formalization}}

We define a domain $d$ as a joint distribution $\mathcal{D}_d(x,y)$ over input-output pairs $(x,y) \in \mathcal{X}\times{}\mathcal{Y}$. Assuming that the training data is sampled from a known mixture of $D$ source domains $\mathcal{D}_s(x,y) = \sum_d \lambda_{s,d} \mathcal{D}_d(x,y)$, several cases are studied in the ML literature:

\begin{itemize}
\item the test distribution $\mathcal{D}_t$ is known 
  \begin{itemize}
  \item $\mathcal{D}_t = \mathcal{D}_{s}$: this is standard ML, with mixture densities
  \item $\mathcal{D}_t(x,y) = \sum_d \lambda_{t,d} \mathcal{D}_d(x,y) $ with a different mixture of domains in training and testing ($\vlambda_s \neq \vlambda_t$), corresponding to multiple source adaptation (MSA) \cite{Mansour09domainadaptation},
  \item $\mathcal{D}_t(x,y) \neq \sum_d \lambda_{t,d} \mathcal{D}_d(x,y) \forall \lambda$, where the target domain for adapation is not a mixture of source domains, studied in \cite{Mansour09multiple}
  \item variants of MSA which only assume knowledge of the marginal distribution $\mathcal{D}_t(x) = \sum_{y} \lambda_{t,d} \mathcal{D}_d(x) $
  \item more realistic situations where  the train and test distributions are estimated from finite samples, also studied in \cite{Mansour09multiple}.
  \end{itemize}
   Note that we recover standard supervised DA is when there is only one component in the training and test mixtures, and multiple-source single domain adaptation when the source is a mixture but the test domain is homogeneous. Unsupervised DA \cite{Daume06domain} corresponds to the case where only the marginal distribution of the test domain inputs is available, while semi-supervised DA is the reverse situation.

\item the test distribution is not known and training needs to be distributionally robust, ie.\ minimize the error against all possible distributions, with variants depending on whether it is or not assumed to be a mixture of existing domains, and whether it is perfectly or imperfectly known. This setting is also studied in eg.\ \cite{Mansour09multiple,Hoffman18algorithms}, and  in \cite{Oren19distributionally} for adversarial language model learning.
\item with respect to the conditional label distribution $P(y|x)$:
  \begin{itemize}
  \item $P(y|x)$ is deterministic, meaning that the joint distribution is entirely specified by the marginal and a labelling function $y=h(x)$; a more complex case is when $h()$ also varies across domains;  
  \item $P(y|x)$ is stochastic, but shared across all domains, which is the typical covariate-shift scenario \cite{Shimodaira00improving}; it can also vary across domains \cite{???}\fyTodo{Missing citation vary accross domains}
  \end{itemize}
\end{itemize}

Multi-domain learning, as defined in \cite{Dredze08online} additionally makes the requirement that categorial domain tags are available in training and testing; a multi-domain learner should use them effectively \cite{Joshi12multidomain} to improve simultaneously on all domains \cite{Finkel09hierarchical}. This is also the view that we adopt here, where we assume training inputs in $(\mathcal{X}\times{}\{1\dots{}D\})$, sampled from a mixture $\sum \lambda_{s,d}\mathcal{D}_{d} \times \{d\}$, and also consider various test scenarios, depending whether $\mathcal{D}_t$ is a fixed mixture of known domains, include new domain(s), in the supervised and unsupervised, and zero-shot settings.\fyTodo{explain zero shot}.

\subsection{MDMT: a tentative definition \label{ssec:definition}}
Informally, a multi-domain system should be able to accurately translate heterogeneous sources of data. This loose definition correspond to several practical situations:
\begin{itemize}
\item the more common situation is when both train and test documents (or sentences) are domain-tagged, where each domain corresponds to a specific source of data, and differ from the other sources in terms of textual genre, thematic content \cite{Zhang16topicinformed}, register, style \cite{Niu18multitask}, etc. It is usually assumed that the set of test domains is the same as the set of training domains -- when this is not the case, we would have zero-shot domain adaptation, similar to what is studied in multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Platanios18contextual}. This is the more general setting, already proposed in \cite{Dredze09multidomain}.\fyTodo{do we need a name here: such as supervised MDMT ? and mostly supervised MDMT for the second case ?}

@@

\subsection{Reasons for using MDMT? \label{ssec:whymdmt}}
The motivations for considering MDMT are many. Looking at the literature, the main arguments in support of MDMT seem to be the following:
\begin{enumerate}
\item[[M1\hspace{-0.5em}]] computational / practical arguments. A first motivation for MDMT is practical  \cite{Sennrich13multidomain,Farajian17neural}: when faced with inputs that are potentially from multiple domains, it is easier and computationally cheaper to develop only one system instead of having to optimize (and maintain) multiple MT engines. Such arguments are all the more convincing that the number of domains considered is large, which is not always the case: for instance \cite{Britz17mixing,Zeng18multidomain,Jiang19multidomain} only consider 2-4 domains in their experiments.

\item[[M2\hspace{-0.5em}]] statistical arguments. The training data available for each domain is very unevenly distributed, and domain trained or adapted on small datasets will yield large estimator variance. Building mix-domain systems is likely to reduce this variance, at the expense of a larger statistical bias \cite{Clark12onesystem}. Under this view, MDMT would be especially useful for domains with little training data. This is what is observed for multilingual MT from English: an increase for under-resource languages due to positive transfer, at the cost of a decrease for well-resourced languages \cite{Arivazhagan19massively}. A slightly distinct line of reasoning is that mixing domains has a positive regularization effect for all domains. By introducing variability in training, it prevents domain adaptation from overfitting the available adaptation data and could help improve generalization even for well-resourced domains. A related case is made in \cite{Joshi12multidomain},
  where the benefits of MD are attributed to an ensembling effect, where several MT systems are simultaneously used in decoding; this effect may subsist even in the absence of clear domain separations.

\item[[M3\hspace{-0.5em}]] linguistic arguments. A third line of reasoning contends that domain specificities are mostly expressed lexically and will primarily affect content words or multi-word expressions; function words, on the other hand, are domain agnostic and remain semantically stable accross domain. A MDMT system should simultaneously learn lexical domain peculiarities, and leverage cross-domain similarities to improve the representation of generic contexts and words \cite{Zeng18multidomain,Pham19generic}. It is here expected that the MDMT scenario should be more profitable when the domain mix includes domains that are closely related and which can share more information. \fyTodo{polysemy}
\end{enumerate}

\fyTodo{recap}
These various motivations for adopting MDMT are not mutually exclusive. Each however yields specific expectations with respect to the performance of this approach: if the motivation is primarily computational, then a drop in MT quality with respect to fine-tuning each domain might be acceptable; if it is to improve statistical estimation, then the hope will be that MDMT will improve, at least for some domains, if not all, over individually trained systems.

% \cite[eq. (3)]{Wang20general} defines the loss of MDMT as the (macro) average of domain loss - all domains are equally important.

@@

\section{Challenging multi-domain translation systems}
In this section, we discuss seven\fyTodo{check number} operational requirements that can be expected from an effective multi-domain system, and discuss ways to evaluate whether these requirements are actually met. % in practice.

\subsection{Multi-domain systems should be effective \label{ssec:effective}}
% \subsubsection{Performance}
The first set of requirements for MDMT systems is that they should perform well in the face of mix-domain data. These yields the following requirements.

\paragraph{[P1-LAB]}\fyTodo{Decide naming scheme} A first expectation for MDMT is that they perform better that the baseline which disregards domain labels, or reassigns them in a random fashion \cite{Joshi12multidomain}.\fyTodo{Try random label assignments ?} % Testing that MDMT actually benefit from domain tags is simple, notwithstanding the previous remarks above regarding evaluation procedures.

\paragraph{[P2-TUN]} Additionally, following motivation~(M2) above, one would expect that MDMT improves over fine-tuning, at least in domains where data is scarce, or in situations where several domains are close. To evaluate this, we perform two measurements, using a real as well as an artificial scenario. In the real scenario, we simply compare the performance of MDMT and fine-tuning for domains of varying sizes, expecting a larger gain for smaller domains (requirement [P2.1-TUN]). In the artificial scenario, we split a single domain in two parts which are considered as distinct in training. The expectation here is that a MDMT should yield a clear gain for both pseudo sub-domains, which should benefit from the supplementary amount of relevant training (requirement [P2.2-TUN]). In this situation, MDMT should even outperform fine-tuning on either of the pseudo sub-domain.
\jcDone{I dont understand the prediction... why running MDMT on 2 artificial subdomains should show  gains over fine-tunning?}\fyTodo{small loss with respect to non split ?}

\subsection{Robustness with respect to domain definition \label{ssec:robusness}}
A second set of requirements is related to the definition of a domain. As repeatedly pointed out in the literature, training corpora for MT are often collected opportunistically and the view that each corpora constitutes a single domain is often a gross approximation.\footnote{Two of our own domains are actually composed of several subcorpora (IT and MED), see details in section~\ref{ssec:corpora}.} MDMT should aim to make the best of the available data and be robust to domain assignments. To challenge these requirements we propose the evaluate the following requirements.

\paragraph{[P3-HET]}
The notion of a domain is often a fragile one and an effective MDMT system should be able to discover not only when cross-domain sharing is useful (cf.\ property [P2.2-TUN])\jcDone{what is P2.2?}, but also when intra-domain heterogeneity is hurting. This requirement is tested by artificially conjoining two separate domains into one during training. The loss in performance with respect to the baseline (correct domain tags) should then be as small as possible.

\paragraph{[P4-ERR]}
A MDMT should perform best when the true domain tag is known, but deteriorate gracefully in the presence of domain tag errors. This is in sharp contrast with what is commonly observed with fine-tuning, where giving the wrong domain tag can yield ``catastrophic'' drop in performance. This requirement is measured by decoding the test set for domain $A$ with all other domain tags, and report average drop in performance.

\paragraph{[P5-UNK]}
A related situation occurs when the domain of a test document is unknown. Several alternatives are in order: (a) automatically assigning one of the domain tag for decoding should not be much worse than using the true domain tag, assuming this domain is seen in training. For test documents from other domains, the same procedure can apply, and should outperform the use of a single mix-domain system.\fyTodo{Systems react to unknown domains}These requirements are respectively labelled [P5.1] and [P5.2]\fyTodo{Not so sure of these ones}.
